Title

### 2024-07-01 paper/main.tex
Udpates

As a result of these efforts, numerous approaches to automated algorithms for
causal discovery have been proposed in the DAG literature, including
constraint-based method like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}. However, the adaption of these algorithms in applied
fields remain limited, and researchers typically prefer to construct DAGs
manually based on their domain expertise \citep{Tennant2020, Petersen2021}.
Even when using automated algorithms, users often rely on manual post hoc
modifications to the model learned by the algorithm because they:

\begin{enumerate}
	\item \textbf{Make Obvious Mistakes:} Although most algorithms are
		theoretically consistent asymptotically, their finite sample
		properties are not well understood. Moreover, depending on the
		selected hyperparameters, their output can vary significantly.
		These issues combined with no good method to test the
		performance of these algorithms on a given dataset, in
		real-world settings with finite samples, these algorithms can
		make obvious mistakes. Consequently, users have to rely on
		posthoc modifications to the model based on their domain
		expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, users have to manually
		orient these edges before using them for analysis.
\end{enumerate}

Due to these challenges, whether constructing the model manually or using an
algorithm, users must rely on their domain expertise to construct or modify the
model. Hence, it is crucial to have tools to guide this modification process.
One of the ways to guide this modification is to use Conditional Independence
(CI) based model testing \cite{Ankan2021}. As DAGs imply CI statements that can
be tested in the given data using statistical CI tests, we can verify whether
the implied CIs are consistent in the data or not. If some of implied CIs are
not valid in the given dataset, we can modify the model accordingly.
Figure~\ref{fig:ci_testing} shows an example of model testing using CIs.
However, this CI testing approach has a few drawbacks. Firstly, as CIs are
implied only by missing edges in the model, we can not test whether an
incorrect edge is already present in the model. Secondly, the number of implied
CIs can be quite large. For example, in the moderately sized alarm network with 
$ 37 $ node, there are $287$ implied CIs. If we rely only on the p-values of the
CI tests, it can be difficult to figure which test to focus on.

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure to show an example of model testing}}
		\label{fig:ci_testing}
\end{figure}
\begin{figure}
	\caption{\todo[inline]{Figure showing an example of modification indices}}
\end{figure}

Inspired by this ranking approach in modification indices, in this paper we
develop a method to rank the CI tests using a conditional measure of
association. Our proposed approach uses the measure of association to find
correlated variables in the data that are not explained by the current model.
Using the measure, we can then rank the pair of variables between which adding
an edge would improve the explanation the most. Based on this information, the
user can select a potential pair of variables and based on their expert
knowledge decide the direction of the edge. Our main contributions in this
paper are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this partial association measure, we present a procedure to
		rank CI tests. This ranking can in turn then be used to
		prioritize addition or deletion of edges from the model.
	\item Lastly, we present a web-browser based tool to allow researchers
		to apply this method to their own datasets.
\end{enumerate}

@@

Here, users have the option to choose the estimators based on the properties of
their data. Non-parametric ensemble estimators such as Random Forest and
XGBoost can work on any combination of data types while performing reasonably
well. For linear datasets users can also choose variants of linear regression
depending on the type of data. After the above steps are repeated for both $ X
$ and $ Y $, we end up with two residual matrices: $ R_X $ and $ R_Y $. The
type of the variable determines the shape for these matrices, i.e., if the
variable is continuous we get a residual matrix of shape $ ( n \times 1 ) $ and
if the variable is categorical we have a residual matrix of shape $ ( n \times
(k-1)) $ (where one of the dummy encoded variables is dropped to avoid
multicollinearity)

We now have two continuous valued residual matrices and the second step now is
to quantify the association between these residuals. In multivariate
statistics, canonical correlation based measures have been used for measuring
association between two sets of variables. We can treat the residuals as sets
of variables in case of categorical variable and use canonical correlations to
quantify the association between them.
\todo[inline]{Justify the argument}

@@

Using this mixed data partial association measure, we propose a method to rank
CI tests. If the user desires to construct the model manually, they can also
start with an empty graph, or start with a DAG outputted by a causal discovery
algorithm. Given a DAG $ G $ and a dataset $ D $, for each pair for variable
$ X $ and $ Y $ in $ G $ that does not have an edge between them, we compute
the partial association measure: $ \rho(X, Y; pa_G(X) \cup pa_G(Y)) $. The
measure of association in this case can be interpreted as the unexplained
observed association between them given the current model. We can then use this
to rank pair of variables which have very high unexplained association between
them. Based on the domain expertise, the user can then choose one of these
suggested pair of variables, decide the direction of the edge between the
variables, and add it to the model. Once the new edge is added we can recompute
the association of all other variables and $ X $ and $ Y $.

@@

\section{Web Tool}
\label{sec:web}
Based on the procedure outlined in the previous section, we built a web-based
tool that allows users to interactively create/modify their model. The user
starts by uploading their dataset and optionally the initial DAG. If no initial
DAG is provided an empty graph is initialized. The tool also computes the
Shipley's C value at any point. The web-tool would automatically show all
unexplained associations in the DAG by adding red edges between them. The edge
weight of the edge represents how high this strength is. Users can also set the
threshold for p-value and association to control how many edges are shown at a
time. Based on these shown potential edges, the user can select the edge that
they want to add and decide the direction for it. If the tool detects an
incorrect edge in the graph, it is shown in black. Once the user is happy with
the graph they have they can export it into different formats to use it for
further computation.

@@

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/plots/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/plots/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC and Hill-Climb Search algorithms against
		manually drawn DAGs using the assistance method. As PC and
		Hill-Climb Search return the Markov Equivalence Class (MEC), we
		use the best and worst scoring orientation of the MEC to get
		the range of values. The human drawn values are done for
	multiple accuracy values of the oracle. \todo[inline]{Replace Hill-Climb with GES}}
\end{figure}

We analyzed the performance of this manual approach by comparing it with two
automated algorithms: PC and Greedy Equivalence Search. For the comparison we
use simulated data from a known DAG and compare how well the algorithm is able
to recover the original model using two metrics: Structural Hamming Distance
(SHD) and Structural Intervention Distance (SID). To simulate the data, we
start with a randomly generated DAG and use linear models with random effects
to generate the data. We repeat the data generation $ 30 $ times for each
density value for the DAG. As both PC and Hill-Climb Search can only give us a
CPDAG, we compare the orientation of these CPDAG which result in the best and
worst case scores on SHD and SID. This procedure is outlined in Algorithm 1.
best and worst case scores on SHD and SID. This procedure is outlined in
Algorithm 1.

\todo[inline]{Add the algorithm}

To simulate an expert using this assisted model construction approach, we start
with an empty model and take a greedy approach to add edges to this model. At
any point, we select the pair of variables that has significant correlation and
the highest unexplained correlation, i.e., highest conditional association. We
then use an oracle to decide the direction of the edge between the variables.
Given an accuracy of the oracle, $ \alpha $, the oracle either gives the
correct direction, incorrect direction, or returns None representing that it
does not know the direction.

@@

If the oracle returns None for any pair of variables, an edge between this pair
is not suggested to the oracle in future iterations. We also make sure that
only edges which do not form a cyclic in the graph are used to select the next
potential edge. After adding each edge, we also check if the p-value of any
existing edge shows independence. If that happens we remove that edge and
blacklist that edge, i.e., that edge is not prompted again during the run of
the algorithm. We repeat this procedure till the model explains all
correlations between the variables. There is a possibility that this greedy
expert gets stuck at incorrect graph structures as it does not do any
backtracking to fix its earlier mistakes. This procedure is outlined in the
Algorithm.
\todo[inline]{Add the algorithm}

\subsection{Results}

>>>>

As a result of these efforts, numerous approaches to automated algorithms for
causal discovery have been proposed in the DAG literature, including
constraint-based method like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}. However, the adaption of these algorithms in applied
fields remain limited, and researchers typically prefer to construct DAGs
manually based on their domain expertise \citep{Tennant2020, Petersen2021}.
Even when using automated algorithms, we often have to rely on manual post hoc
modifications to the model learned by the algorithm because they:

\begin{enumerate}
	\item \textbf{Make Obvious Mistakes:} Although most algorithms are
		theoretically consistent asymptotically, their finite sample
		properties are not well understood. Moreover, depending on the
		selected hyperparameters, their output can vary significantly.
		These issues combined with no good method to test the
		performance of these algorithms on a given dataset, in
		real-world settings with finite samples, these algorithms can
		make obvious mistakes. Consequently, researchers have to rely on
		posthoc modifications to the model based on their domain
		expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

Due to these challenges, whether constructing the model manually or using an
algorithm, we must rely on domain expertise to construct or modify the
model. Hence, it is crucial to have tools to guide this modification process.
One of the ways to guide this modification is to use Conditional Independence
(CI) based model testing \cite{Ankan2021}. As DAGs imply CI statements that can
be tested in the given data using statistical CI tests, we can verify whether
the implied CIs are consistent in the data or not. If some of implied CIs are
not valid in the given dataset, we can modify the model accordingly.
Figure~\ref{fig:ci_testing} shows an example of model testing using CIs.
However, this CI testing approach has a few drawbacks. Firstly, as CIs are
implied only by missing edges in the model, we can not test whether an
incorrect edge is already present in the model. Secondly, the number of implied
CIs can be quite large. For example, in the moderately sized alarm network with 
$ 37 $ node, there are $287$ implied CIs. If we rely only on the p-values of the
CI tests, it can be difficult to figure which test to focus on.

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure. Show an example of model testing using CIs}}
		\label{fig:ci_testing}
\end{figure}
\begin{figure}
	\caption{\todo[inline]{Figure showing an example of modification indices}}
\end{figure}

Inspired by this ranking approach in modification indices, in this paper we
develop a method to rank the CI tests using a conditional measure of
association. Our proposed approach uses the measure of association to find
correlated variables in the data that are not explained by the current model.
Using the measure, we can then rank the pair of variables between which adding
an edge would improve the explanation the most. Based on this information, the
user can select a potential pair of variables and based on their expert
knowledge decide the direction of the edge. Our main contributions in this
paper are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this partial association measure, we present a procedure to
		rank CI tests. This ranking can be used to prioritize addition
		or deletion of edges from the model.
	\item Lastly, we present a web-browser based tool to allow researchers
		to apply this method to their own datasets.
\end{enumerate}

@@

Here, we have the option to choose the estimators based on the properties of
their data. Non-parametric ensemble estimators such as Random Forest and
XGBoost can work on any combination of data types while performing reasonably
well. For linear datasets we can also choose variants of linear regression
depending on the type of data. After the above steps are repeated for both $ X
$ and $ Y $, we end up with two residual matrices: $ R_X $ and $ R_Y $. The
type of the variable determines the shape for these matrices, i.e., if the
variable is continuous we get a residual matrix of shape $ ( n \times 1 ) $ and
if the variable is categorical we have a residual matrix of shape $ ( n \times
(k-1)) $ (where one of the dummy encoded variables is dropped to avoid
multicollinearity)

We now have two continuous valued residual matrices and the second step now is
to quantify the association between these residuals. In multivariate
statistics, canonical correlation based measures have been used for measuring
association between two sets of variables. We can treat the residuals as sets
of variables in case of categorical variable and use canonical correlations to
quantify the association between them.

@@

Using this mixed data partial association measure, we propose a method to rank
CI tests, and use it for model construction and modification. If the user
desires to construct the model manually, they can also start with an empty
graph, or start with a DAG outputted by a causal discovery algorithm. Given a
DAG $ G $ and a dataset $ D $, for each pair for variable $ X $ and $ Y $ in $
G $ that does not have an edge between them, we compute the partial association
measure: $ \rho(X, Y; pa_G(X) \cup pa_G(Y)) $. The measure of association in
this case can be interpreted as the unexplained observed association between
them given the current model. We can then use this to rank pair of variables
which have very high unexplained association between them. Based on the domain
expertise, we can then choose one of these suggested pair of variables, decide
the direction of the edge between the variables, and add it to the model. Once
the new edge is added we can recompute the association of all other variables
and $ X $ and $ Y $.

@@

\section{Web Tool}
\label{sec:web}
Based on the procedure outlined in the previous section, we built a web-based
tool that allows us to interactively create/modify their model. The user
starts by uploading their dataset and optionally the initial DAG. If no initial
DAG is provided an empty graph is initialized. The tool also computes the
Shipley's C value at any point. The web-tool would automatically show all
unexplained associations in the DAG by adding red edges between them. The edge
weight of the edge represents how high this strength is. Users can also set the
threshold for p-value and association to control how many edges are shown at a
time. Based on these shown potential edges, the user can select the edge that
they want to add and decide the direction for it. If the tool detects an
incorrect edge in the graph, it is shown in black. Once the user is happy with
the graph they have they can export it into different formats to use it for
further computation.

@@

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/plots/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/plots/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC and Hill-Climb Search algorithms against
		manually drawn DAGs using the assistance method. As PC and
		Hill-Climb Search return the Markov Equivalence Class (MEC), we
		use the best and worst scoring orientation of the MEC to get
		the range of values. The human drawn values are done for
	multiple accuracy values of the oracle. \todo[inline]{Replace Hill-Climb with GES and improve the plot}}
\end{figure}

We analyzed the performance of this manual approach by comparing it with two
automated algorithms: PC and Greedy Equivalence Search (GES). For the
comparison we use simulated data from a known DAG and compare how well the
algorithm is able to recover the original model using two metrics: Structural
Hamming Distance (SHD) and Structural Intervention Distance (SID). To simulate
the data, we start with a randomly generated DAG on $ 10 $ nodes and use linear
models with random effects to generate the data.

We vary the edge probability of the DAG to generate DAGs with varying density.
We repeat the data generation $ 30 $ times for each density value for the DAG.
As both PC and Hill-Climb Search can only give us a CPDAG, we compare the
orientation of these CPDAG which result in the best and worst case scores on
SHD and SID.

To use our proposed approach, we need an expert to tell us the direction of the
edges. We simulated an expert using this assisted model construction approach,
we start with an empty model and take a greedy approach to add edges to this
model. At any point, we select the pair of variables that has significant
correlation and the highest unexplained correlation, i.e., highest conditional
association. We then use an oracle to decide the direction of the edge between
the variables. Given an accuracy of the oracle, $ \alpha $, the oracle either
gives the correct direction, incorrect direction, or returns None representing
that it does not know the direction.

@@

If the oracle returns None for any pair of variables, an edge between this pair
is not suggested to the oracle in future iterations. We also make sure that
only edges which do not form a cyclic in the graph are used to select the next
potential edge. After adding each edge, we also check if the p-value of any
existing edge shows independence. If that happens we remove that edge and
blacklist that edge, i.e., that edge is not prompted again during the run of
the algorithm. We repeat this procedure till the model explains all
correlations between the variables. There is a possibility that this greedy
expert gets stuck at incorrect graph structures as it does not do any
backtracking to fix its earlier mistakes. This procedure is outlined in the
Algorithm 1.
\todo[inline]{Show an algorithm of how this expert method is working}
